---
title: "Analysing malariasimulation output with postie"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing malariasimulation output with postie}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Postie provides tools to post-process output from
[malariasimulation](https://mrc-ide.github.io/malariasimulation/).
This vignette runs a short example simulation and explores how to
summarise the resulting prevalence and incidence estimates.

## Running a basic simulation

```{r}
# install.packages("malariasimulation")
library(malariasimulation)
library(postie)
library(dplyr)
library(ggplot2)

# set up parameters
year <- 365
p <- get_parameters() |>
  set_drugs(list(AL_params)) |>
  set_clinical_treatment(1, 0.5, 0.5) |>
  set_epi_outputs(
    age_group = c(0, 5, 15, 200) * year,
    prevalence = c(2, 10) * year,
    clinical_incidence = c(0, 5, 15, 200) * year,
    severe_incidence = c(0, 5, 15, 200) * year
  )

# run for two years
sim <- run_simulation(2 * year, p)
```

The object `sim` is a data frame containing the time series of all
requested outputs. Postie converts these counts into rates that are
easier to interpret.

## Rates and prevalence

```{r}
rates <- get_rates(sim)
#> Warning: required column `ft_sev` (probability hospitalisation | severe case) not found, assuming ft_sev = 0.8
head(rates)

sim <- dplyr::mutate(sim, ft_sev = 0.5)
rates <- get_rates(sim)

prev <- get_prevalence(sim)
head(prev)
```

### Handling `ft_sev`

The column `ft_sev` contains the probability that a severe case leads to
hospital admission. When it is missing `get_rates()` assumes a value of
0.8 and prints a warning. By supplying your own column before calling the
function you can control this assumption and remove the warning.

For the full methodology behind how postie adjusts severe incidence and
derives mortality from `ft_sev` (and the related `treatment_scaler`,
`hosp_sev_cfr` and `community_sev_cfr` arguments), see the
`vignette("severe-mortality")`.

The rates are per person per day. Multiplying by `365` converts them to
the more familiar per person per year units.



```{r}
prev |>
  ggplot(aes(time, lm_prevalence_2_10)) +
  theme_bw() +
  geom_line() +
  labs(y = expression(PfPr[2-10]))
```

## Units and time aggregation

Rates returned by `get_rates()` are expressed per person per day. The unit
is separate from the time interval over which values are aggregated. You
can therefore average rates by month or year while still reporting them as
cases per person per year by multiplying by `365`.

## Aggregating over time and age

The flexibility of postie means you can aggregate outputs in many ways.
For example, to obtain annual clinical incidence and mortality by the
three age groups:

```{r}
summary <- rates |>
  mutate(
    age_group = case_when(
      age_upper <= 5 ~ "U5",
      age_upper <= 15 ~ "5-15",
      TRUE ~ "15+"
    )
  ) |>
  summarise(
    clinical = weighted.mean(clinical, person_days) * 365,
    mortality = weighted.mean(mortality, person_days) * 365,
    person_days = sum(person_days),
    time = mean(time),
    .by = c(year, age_group)
  )
summary
```

Aggregating by both year and month can reveal seasonal patterns. The
following code collapses the daily values to monthly rates and plots the
clinical incidence over time.

```{r}
monthly <- rates |>
  summarise(
    clinical = weighted.mean(clinical, person_days) * 365,
    time = mean(time),
    .by = c(year, month)
  )

monthly |>
  ggplot(aes(time, clinical)) +
  theme_bw() +
  geom_line() +
  labs(y = "clinical per person per year")
```

Postie therefore provides a straightforward pipeline from raw model output to tidy, aggregated indicators ready for further analysis or visualisation.

